`r meta$title`

### 2022-12-14 paper.Rmd
reference  setups for specific outlets to accommodate differences in non-numbered/numbered sections and output types (PDF/HTML)

---
title: Current state and prospects of R-packages for the design of experiments
author:
  - name: Emi Tanaka
    affiliation: Monash University
  - name: Dewi Amaliah 
    affiliation: Monash University
abstract: >
    Re-running an experiment is generally costly and, in some cases, impossible due to limited resources; therefore, the design of an experiment plays a critical role in increasing the quality of experimental data. In this paper, we describe the current state of R-packages for the design of experiments through an exploratory data analysis of package downloads, package metadata, and a comparison of characteristics with other topics. We observed that experimental designs in practice appear to be sufficiently manufactured by a small number of packages, and the development of experimental designs often occurs in silos. We also discuss the interface designs of widely utilized R packages in the field of experimental design and discuss their future prospects for advancing the field in practice.
bibliography: paper.bib
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) })
keywords: ["experimental design", "CRAN task view", "interface design"]
output: 
  #word_document: default
  pagedown::html_paged:
    template: assets/html-template.html
    csl: assets/html-template.csl
    self_contained: true
    css: ["assets/html-template.css"]
---

```{r setup, cache = FALSE, include = FALSE}
library(glue)
library(ggwordcloud)
library(tidyverse)
library(patchwork)
library(tidygraph)
library(ggraph)
theme_set(
  theme(
    panel.background = element_rect(fill = NA),
    panel.grid = element_line(color = "lightgray"),
    axis.text = element_text(color = "black"),
    axis.line = element_line(color = "black", size = 0.7),
    axis.ticks.length = unit(1.4, "mm"),
    axis.ticks = element_line(color = "black", size = 0.7),
    axis.title = element_text(color = "black", face = "bold"),
    strip.background = element_rect(color = "black",
                                    fill = "white"),
    strip.text = element_text(color = "black",
                              margin = margin(5, 5, 5, 5)),
    plot.title.position = "plot",
    plot.title = element_text(color = "black", hjust = 0)
  )
)
knitr::opts_chunk$set(echo = FALSE, 
                      cache = TRUE,
                      cache.path = "cache/",
                      fig.align = 'center', 
                      fig.pos = 'htbp', 
                      fig.path = "figures/",
                      fig.width = 6,
                      fig.height = 6.5,
                      message = FALSE,
                      warning = FALSE)

ctv <- function(view) {
  ifelse(knitr::is_html_output(), glue("[{view}](https://cran.r-project.org/web/views/{view}.html)"), glue("\\ctv{[view]}", .open = "[", .close = "]"))
}
CRANpkg <- function(pkg) {
  ifelse(knitr::is_html_output(), glue("[{pkg}](https://cran.r-project.org/web/packages/{pkg}/index.html)"), glue("\\CRANpkg{[pkg]}", .open = "[", .close = "]"))
}
ref <- function(x) {
  case_when(knitr::is_html_output() ~ glue("\\@ref({x})"),
            TRUE ~ glue("\\ref{[x]}", .open = "[", .close = "]"))
}
url <- function(x) {
  ifelse(knitr::is_html_output(), x, glue("\\url{[x]}", .open = "[", .close = "]"))
}
rpkg <- function(x) {
  glue("`{x}`")
}

ctv_pkg_data <- readRDS("output/ctv_pkg_data.rds")
ctv_summary_table <- readRDS("output/ctv_summary_table.rds")
doe_bigram_all_with_download_data <- readRDS("output/doe_bigram_all_with_download_data.rds")
doe_pkg_network <- readRDS("output/doe_pkg_network.rds")
ctv_gini_yearly <- readRDS("output/ctv_gini_yearly.rds")
ctv_lorenz_yearly <- readRDS("output/ctv_lorenz_yearly.rds")
ctv_rank_yearly <- readRDS("output/ctv_rank_yearly.rds")
access_date_pkg_downloads <- readRDS("output/access_date_pkg_downloads.rds")

@@

The utility of the software can also be described by its design to facilitate the clear expression and interpretation of the desired experimental design. Certain programming language designs can hinder or discourage the development of reliable programs [@Wasserman1975-xr].  The immense popularity of `r rpkg("tidyverse")` [a collection of R-packages for various stages of data analysis that places enormous emphasis on the interface design by @Wickham2019-mj] is a testament to the impact that an interface design can have in practice. The practice of experimental design can be advanced by adopting similar interface design principles across the DoE packages.

The remainder of this paper is organized as follows. Section `r ref("data")` briefly describes the data source used for the analysis in Section `r ref("eda")`; Section `r ref("eda")` presents some insights into the state of the current DoE packages by the exploratory data analysis of package download data, text descriptions and comparisons with other CRAN task views; Section `r ref("design")` discusses the interface designs of widely used DoE packages, and we conclude with a discussion in Section `r ref("discussion")` of future prospects in the software development of experimental designs.

@@

The Comprehensive R Archive Network (CRAN) is a network of servers located across the world that stores mirrored versions of the R and R packages. The most popular network is the RStudio mirror (the default server for those that use the RStudio IDE). The RStudio mirror is also the only server that provides a comprehensive daily download logs of R and R packages since October 2012. The summary data can be easily accessed using the `r rpkg("cranlogs")` package [@cranlogs]. This paper uses the data from the beginning of 2013 to the end of 2021 (a total of nine years) for the packages in the CRAN task views.

@@

All CRAN packages have a title, description, package connections (suggests, depends, and imports of other packages), and other meta-information in the DESCRIPTION file. We use text data from the title and description (accessed in `r access_date_pkg_downloads`) in Section `r ref("topics")`.

@@

CRAN task views are volunteer-maintained lists of R-packages on CRAN relevant to the corresponding topic. There were `r n_distinct(ctv_pkg_data$name)` CRAN task views in total. Table S1 in the Supplementary Materials list the available topics from the `r rpkg("ctv")` package [@ctv].
The list of packages in each CRAN task view (as of `r access_date_pkg_downloads`) is used to contrast the characteristics of the DoE packages in Sections `r ref("popular")` and `r ref("silo")`.

# Explorative data analysis {#eda}

In this section, we derive some conjectures based on an analysis of the data described in Section `r ref("data")`. All results presented are from exploratory data analysis of observational data, consequently, all interpretations are somewhat speculative and may not be indicative of the true state of the field of experimental design. In particular, any analysis over time is confounded by the fact that the nature of users and package management has changed over the years. It should be noted that some DoE packages may have been archived or removed from the task view over the years; therefore, any cross-sectional analysis presented may not reflect the set of all DoE packages at that particular time period (although we assume such incidences are low).

@@

## Small, but diverse, set of packages are sufficient for most experimental designs in practice {#popular}

@@

An increase in the number of packages that are not highly downloaded may mean that **_there are more packages to construct niche experimental designs_**. Some examples of these packages include `r rpkg("qtlDesign")`, `r rpkg("PwrGSD")` and `r rpkg("Crossover")` made for QTL experiments, group sequential designs and crossover trials, respectively. These packages would naturally have fewer potential users. Counterfactual to this, the increase could be due to other external factors, such as an increase in the number of skilled developers (and thus more package contributions), a change in CRAN policy or management to add packages (either to CRAN and/or task view), and/or the fact that new packages are still yet to amass users. While there is an argument that low download counts are due to the low utility and/or quality of packages, packages in CRAN task views are selected by expert maintainers. We can reasonably assume that any package listed in the CRAN task view has an acceptable utility and quality.

@@

## The field of experimental design is slow-changing  {#ranking}

We can see in Figure `r ref("fig:rank-over-time")` that most of the top 10 ranking packages have been in the top 10 for the last nine years with `r rpkg("lhs")` steadily climbing up the ranks in the last few years. It should be noted that the download of one package can prompt the download of another package; the most notable package connection is `r rpkg("AlgDesign")` and `r rpkg("agricolae")`, where the former is an import for the latter. The full network of package connections within the DoE packages is shown in Figure `r ref("fig:plot-doe-network")`.

@@

## Optimal designs are of interest {#topics}

@@

1. We standardized the words to the lower case and removed pluralization. 
2. Multiple mentions of the same bigram within a package were counted as one (for example, `r rpkg("AlgDesign")` mentions "experimental design" four times in the title and the description, but this is counted as one). 
3. Bigrams consisting of stop words were removed. The stop words are sourced from the lexicons in `tidytext::stop_words` in addition to other words we deemed irrelevant, e.g. "provide", "e.g.", "calculate" and so on -- full list is shown in the code provided in the link under Section `r ref("pkgs")`.

@@

## Development of experimental designs occur in silos {#silo}

@@

In this section, we discuss the interface designs of functions that output an experimental design based on three broad areas: factorial, recipe and augmenting designs. The discussion is exclusive to the top downloaded packages (shown in Figure `r ref("fig:rank-over-time")`), with the exception of `r rpkg("ez")` and `r rpkg("DiceKriging")`, as the former is predominately visualization of experimental data and the latter is about the analysis of computer experiments in addition to belonging to the same suite of packages as `r rpkg("DiceDesign")` [@DiceDesign].

@@

The `r rpkg("DoE.base")` package [@DoE.base] can construct full factorial and (regular and irregular) orthogonal array designs via `fac.design()` and `oa.design()`, respectively. The `r rpkg("FrF2")` package [@FrF2] constructs regular fractional 2-level factorial designs using `FrF2()` and `FrF2Large()`, with the latter for large designs.  The `r rpkg("conf.design")` package  [@conf.design] constructs symmetric confounded factorial designs via `conf.design()` and the `r rpkg("BHH2")` package [@BHH2] generates a full or fractional 2-level factorial design matrix via `ffDesMatrix()`.

@@

The output of the design is either a special class of list (e.g. the `design` class for `r rpkg("DoE.base")` and `r rpkg("FrF2")` or a `data.frame` for `r rpkg("conf.design")`) or a matrix for `BHH2` such that an element or column corresponds to a treatment factor with each value corresponding to one experimental run. The treatment factors are generally assigned pseudo names (e.g. letters of the alphabet) or an argument exists for users to input treatment names as a character vector.

Some forms of factorial design are known by other names, for example, *response surface designs* are factorial designs where the treatment factors are discrete levels of continuous variables. Two types of response surface designs can be constructed using the `r rpkg("rsm")` package [@rsm]: Box-Behnken design [@Box1960-wr] and central-composite designs [@Box1951-ji] via functions `bbd()` and `ccd()`, respectively, where the minimum required input is the number of factors. Another form of factorial design is the *saturated designs* where higher-order interaction effects of treatment factors are typically confounded with the main effects. Plankett-Burnman designs [@Plackett1946-ly] are a type of saturated design that can be generated by the function `pb()` in the `r rpkg("FrF2")` package, where the user provides the number of experimental runs and the number of treatment factors.

@@

The `r rpkg("agricolae")` package [@agricolae] is the prime example of constructing designs based on a set of so-called "recipe functions", where each function corresponds to a single class of experimental design. For example, `design.crd()`, `design.rcbd()` and `design.split()` construct completely randomized, randomized complete block and split-plot designs, respectively. Users typically supply treatment labels (or the number of treatments in the case of `design.split()`) and the number of replications as arguments for these functions. The output is a list with one element corresponding to a `data.frame` that contains the design in a table such that the row corresponds to the experimental run and the columns correspond to the experimental variables (we refer to this format simply as "table format" henceforth).

The use of recipe functions is not limited to classical experimental designs; the `r rpkg("AlgDesign")` [@AlgDesign] package offers three primary functions for generating optimal designs: `optBlock()`, `optFederov()` and `optMonteCarlo()`. In general, these functions require data and formulas in terms of the supplied data variables with the choice of the criterion (e.g. the D-criterion), with the output as a list with one element corresponding to the design in a table format. The difference between these functions lies in the underlying search strategy for optimal designs, and the name of the function is a surrogate for the search algorithm.

Computer experiments, which generally involve space-filling designs, are implemented in packages such as `r rpkg("lhs")` [@lhs] and `r rpkg("DiceDesign")` [@DiceDesign]. For the `r rpkg("lhs")` package, functions such as `randomLHS()`, `optimumLHS()`, and `maximinLHS()`, require users to specify the sample size ($n$) and the number of variables ($p$), then it generates a Latin hypercube sample [@McKay1979-aw], based on different optimization schemes (in this case, random, S optimal and maxmin criteria, respectively; see package documentation for more details). Similarly for `r rpkg("DiceDesign")`, there is a comprehensive list of space-filling designs such as `dmaxDesign()`, `lhsDesign()`, and `wspDesign()` with input of $n$ and $p$ as before (among additional parameters for some) that implement algorithms that either maximize the entropy [@Shewry1987-jn], produce a random Latin hypercube sample, or use the WSP algorithm [@Santiago2012-ui], respectively. These designs output an $n \times p$ matrix with values between 0 and 1. Again, these functions employ a recipe style, in which each function has a name that corresponds to a certain search strategy to generate the experimental design.

@@

Some functions in the DoE packages require the input of existing experimental designs to produce new designs. For example, the `r rpkg("DoE.base")` package contains some experimental functions, `cross.design()` and `param.design()`, to combine designs with the former taking a Cartesian product of the input designs while the latter uses a Taguchi style [@Taguchi1986-ut] to aggregate the designs with the inner and outer arrays. The `r rpkg("lhs")` package contains a function that `augmentLHS()` to add additional samples to the existing Latin hypercube sample.

Another class of augmenting design is *sequential design* (also called *adaptive sampling*), which is best represented by `r rpkg("tgp")` [@tgp]. This requires prior information that is used to inform the next experimental design using `tgp.design()` and `dopt.gp()`. The user is required to supply candidate samples to subsample from and a model or a prior experimental design. Follow-up experiments, which can also be classified as sequential designs, is implemented by `r rpkg("BsMD")` [@BsMD] using a model-discriminant approach using `MD()`.

@@

Through the exploratory data analysis of the three data sources (package download logs, package metadata and CRAN task views) outlined in Section `r ref("data")`, we observed in Section `r ref("eda")` that the the total download of DoE packages is concentrated only on a handful of R-packages although these represent a diverse set in comparison with other CRAN task views. Furthermore, the data suggest that experimental design is the least collaborative field.

@@

The interface design (discussed in Section `r ref("design")`) reveals that the most widely used DoE packages generally have functions that 1) focus on certain aspects of experimental design (e.g. factorial structure or augmenting design), 2) are a recipe format (i.e. the name of the function is a surrogate to a single class of the design or optimal search algorithm) and 3) context is often a second thought -- many inputs a single integer corresponding to the number of factors, levels, or experimental runs (or sample size). The function will often assign pseudo-factor names or there is an optional argument to input a character vector that corresponds to the factor names. These interface designs require users to have processed the experiment in statistical terms (often stripping the experimental context away) and simultaneously, users must choose the generation mechanism by selecting an appropriate function. Arguably, the current dominant interface designs are not aligned with the way practitioners cognitively design their experiments. Often, the critical part of designing an experiment is to understand the experimental structure, and the experimental context can govern or guide the choice of algorithm to allocate treatments. In addition, the nature of recipe functions can obscure the understanding and relation of designs (e.g., how do you go from an unstructured factorial design to split plot design?). Each new method for generating an experimental design appears to correspond to a completely new function, and intermediate results are often not easily accessible. These factors may contribute to why developers often work in silos in the field of experimental design.

A consistent cognitive interface design that leverages existing developments and can be easily extended to new methods will conceivably be of great help to practitioners. Some efforts to this end are seen in `r rpkg("DoE.wrapper")` [@DoE.wrapper], which contains wrapper recipe functions for other DoE packages such as `r rpkg("lhs")`, `r rpkg("AlgDesign")` and `r rpkg("FrF2")` (see Figure `r ref("fig:plot-doe-network")`), and is also the subject of the developmental package `r rpkg("edibble")` [@edibble]. Undoubtedly, no single developer or package can cater to all experimental designs; therefore, any unifying interface should consider how other developers can contribute or add their methods. Future research could benefit from further exploratory data analysis, expanding the study to beyond R-packages, and discussing other aspects of interface designs.

@@

This paper uses `r rpkg("targets")` [@targets] and `r rpkg("renv")` [@renv] for reproducibility, `r rpkg("knitr")` [@knitr] and `r rpkg("rmarkdown")` [@rmarkdown] for creating reproducible documents, `r rpkg("ggplot2")` [@ggplot2], `r rpkg("ggraph")` [@ggraph], `r rpkg("ggwordcloud")` [@ggwordcloud] and `colorspace` [@colorspace] for visualization, `r rpkg("kableExtra")` [@kableExtra] for customizing the table in the Supplementary Material and `r rpkg("tidyverse")` [@Wickham2019-mj], `r rpkg("tidytext")` [@tidytext], `r rpkg("pluralize")` [@pluralize], and `r rpkg("ineq")` [@ineq] for data processing and manipulation, and `r rpkg("cranlogs")` [@cranlogs] and `r rpkg("ctv")` [@ctv] for extracting data. All codes used to reproduce this paper are found in `r url("https://github.com/emitanaka/paper-DoE-review")`.

>>>>

---
title: Current state and prospects of R-packages for the design of experiments
author:
  - name: Emi Tanaka
    affiliation: Monash University
  - name: Dewi Amaliah 
    affiliation: Monash University
abstract: >
    Re-running an experiment is generally costly and, in some cases, impossible due to limited resources; therefore, the design of an experiment plays a critical role in increasing the quality of experimental data. In this paper, we describe the current state of R-packages for the design of experiments through an exploratory data analysis of package downloads, package metadata, and a comparison of characteristics with other topics. We observed that experimental designs in practice appear to be sufficiently manufactured by a small number of packages, and the development of experimental designs often occurs in silos. We also discuss the interface designs of widely utilized R packages in the field of experimental design and discuss their future prospects for advancing the field in practice.
bibliography: paper.bib
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) })
keywords: ["experimental design", "CRAN task view", "interface design"]
output: 
  #word_document: default
  pagedown::html_paged:
    template: assets/html-template.html
    csl: assets/html-template.csl
    self_contained: true
    css: ["assets/html-template.css"]
params:
  ref_setup_fn: "ref-setup.R"
---

```{r setup, cache = FALSE, include = FALSE}
library(glue)
library(ggwordcloud)
library(tidyverse)
library(patchwork)
library(tidygraph)
library(ggraph)
theme_set(
  theme(
    panel.background = element_rect(fill = NA),
    panel.grid = element_line(color = "lightgray"),
    axis.text = element_text(color = "black"),
    axis.line = element_line(color = "black", size = 0.7),
    axis.ticks.length = unit(1.4, "mm"),
    axis.ticks = element_line(color = "black", size = 0.7),
    axis.title = element_text(color = "black", face = "bold"),
    strip.background = element_rect(color = "black",
                                    fill = "white"),
    strip.text = element_text(color = "black",
                              margin = margin(5, 5, 5, 5)),
    plot.title.position = "plot",
    plot.title = element_text(color = "black", hjust = 0)
  )
)
knitr::opts_chunk$set(echo = FALSE, 
                      cache = TRUE,
                      cache.path = "cache/",
                      fig.align = 'center', 
                      fig.pos = 'htbp', 
                      fig.path = "figures/",
                      fig.width = 6,
                      fig.height = 6.5,
                      message = FALSE,
                      warning = FALSE)
source(params$ref_setup_fn)
```

```{r data, include = FALSE}
ctv_pkg_data <- readRDS("output/ctv_pkg_data.rds")
ctv_summary_table <- readRDS("output/ctv_summary_table.rds")
doe_bigram_all_with_download_data <- readRDS("output/doe_bigram_all_with_download_data.rds")
doe_pkg_network <- readRDS("output/doe_pkg_network.rds")
ctv_gini_yearly <- readRDS("output/ctv_gini_yearly.rds")
ctv_lorenz_yearly <- readRDS("output/ctv_lorenz_yearly.rds")
ctv_rank_yearly <- readRDS("output/ctv_rank_yearly.rds")
access_date_pkg_downloads <- readRDS("output/access_date_pkg_downloads.rds")

@@

The utility of the software can also be described by its design to facilitate the clear expression and interpretation of the desired experimental design. Certain programming language designs can hinder or discourage the development of reliable programs [@Wasserman1975-xr].  The immense popularity of `r CRANpkg("tidyverse")` [a collection of R-packages for various stages of data analysis that places enormous emphasis on the interface design by @Wickham2019-mj] is a testament to the impact that an interface design can have in practice. The practice of experimental design can be advanced by adopting similar interface design principles across the DoE packages.

The remainder of this paper is organized as follows. `r secref("data", "data")` briefly describes the data source used for the analysis; `r secref("eda", "Exploratory data analysis" )` presents some insights into the state of the current DoE packages by the exploratory data analysis of package download data, text descriptions and comparisons with other CRAN task views; `r secref("design", "Interface design")` discusses the interface designs of widely used DoE packages, and we conclude with a discussion in `r secref("discussion", "Discussion")` of future prospects in the software development of experimental designs.

@@

The Comprehensive R Archive Network (CRAN) is a network of servers located across the world that stores mirrored versions of the R and R packages. The most popular network is the RStudio mirror (the default server for those that use the RStudio IDE). The RStudio mirror is also the only server that provides a comprehensive daily download logs of R and R packages since October 2012. The summary data can be easily accessed using the `r CRANpkg("cranlogs")` package [@cranlogs]. This paper uses the data from the beginning of 2013 to the end of 2021 (a total of nine years) for the packages in the CRAN task views.

@@

All CRAN packages have a title, description, package connections (suggests, depends, and imports of other packages), and other meta-information in the DESCRIPTION file. We use text data from the title and description (accessed in `r access_date_pkg_downloads`).

@@

CRAN task views are volunteer-maintained lists of R-packages on CRAN relevant to the corresponding topic. There were `r n_distinct(ctv_pkg_data$name)` CRAN task views in total. Table S1 in the Supplementary Materials list the available topics from the `r CRANpkg("ctv")` package [@ctv].
The list of packages in each CRAN task view (as of `r access_date_pkg_downloads`) is used to contrast the characteristics of the DoE packages.

# Exploratory data analysis {#eda}

In this section, we derive some conjectures based on an analysis of the data described in `r secref("data", "Data")`. All results presented are from exploratory data analysis of observational data, consequently, all interpretations are somewhat speculative and may not be indicative of the true state of the field of experimental design. In particular, any analysis over time is confounded by the fact that the nature of users and package management has changed over the years. It should be noted that some DoE packages may have been archived or removed from the task view over the years; therefore, any cross-sectional analysis presented may not reflect the set of all DoE packages at that particular time period (although we assume such incidences are low).

@@

## Small, but diverse, set of packages are sufficient for most experimental designs in practice

@@

An increase in the number of packages that are not highly downloaded may mean that **_there are more packages to construct niche experimental designs_**. Some examples of these packages include `r CRANpkg("qtlDesign")`, `r CRANpkg("PwrGSD")` and `r CRANpkg("Crossover")` made for QTL experiments, group sequential designs and crossover trials, respectively. These packages would naturally have fewer potential users. Counterfactual to this, the increase could be due to other external factors, such as an increase in the number of skilled developers (and thus more package contributions), a change in CRAN policy or management to add packages (either to CRAN and/or task view), and/or the fact that new packages are still yet to amass users. While there is an argument that low download counts are due to the low utility and/or quality of packages, packages in CRAN task views are selected by expert maintainers. We can reasonably assume that any package listed in the CRAN task view has an acceptable utility and quality.

@@

## The field of experimental design is slow-changing

We can see in Figure `r ref("fig:rank-over-time")` that most of the top 10 ranking packages have been in the top 10 for the last nine years with `r CRANpkg("lhs")` steadily climbing up the ranks in the last few years. It should be noted that the download of one package can prompt the download of another package; the most notable package connection is `r CRANpkg("AlgDesign")` and `r CRANpkg("agricolae")`, where the former is an import for the latter. The full network of package connections within the DoE packages is shown in Figure `r ref("fig:plot-doe-network")`.

@@

## Optimal designs are of interest

@@

1. We standardized the words to the lower case and removed pluralization. 
2. Multiple mentions of the same bigram within a package were counted as one (for example, `r CRANpkg("AlgDesign")` mentions "experimental design" four times in the title and the description, but this is counted as one). 
3. Bigrams consisting of stop words were removed. The stop words are sourced from the lexicons in `tidytext::stop_words` in addition to other words we deemed irrelevant, e.g. "provide", "e.g.", "calculate" and so on -- full list is shown in the code provided in the link under `r secref("pkgs", "Acknowledgement")`.

@@

## Development of experimental designs occur in silos

@@

In this section, we discuss the interface designs of functions that output an experimental design based on three broad areas: factorial, recipe and augmenting designs. The discussion is exclusive to the top downloaded packages (shown in Figure `r ref("fig:rank-over-time")`), with the exception of `r CRANpkg("ez")` and `r CRANpkg("DiceKriging")`, as the former is predominately visualization of experimental data and the latter is about the analysis of computer experiments in addition to belonging to the same suite of packages as `r CRANpkg("DiceDesign")` [@DiceDesign].

@@

The `r CRANpkg("DoE.base")` package [@DoE.base] can construct full factorial and (regular and irregular) orthogonal array designs via `fac.design()` and `oa.design()`, respectively. The `r CRANpkg("FrF2")` package [@FrF2] constructs regular fractional 2-level factorial designs using `FrF2()` and `FrF2Large()`, with the latter for large designs.  The `r CRANpkg("conf.design")` package  [@conf.design] constructs symmetric confounded factorial designs via `conf.design()` and the `r CRANpkg("BHH2")` package [@BHH2] generates a full or fractional 2-level factorial design matrix via `ffDesMatrix()`.

@@

The output of the design is either a special class of list (e.g. the `design` class for `r CRANpkg("DoE.base")` and `r CRANpkg("FrF2")` or a `data.frame` for `r CRANpkg("conf.design")`) or a matrix for `BHH2` such that an element or column corresponds to a treatment factor with each value corresponding to one experimental run. The treatment factors are generally assigned pseudo names (e.g. letters of the alphabet) or an argument exists for users to input treatment names as a character vector.

Some forms of factorial design are known by other names, for example, *response surface designs* are factorial designs where the treatment factors are discrete levels of continuous variables. Two types of response surface designs can be constructed using the `r CRANpkg("rsm")` package [@rsm]: Box-Behnken design [@Box1960-wr] and central-composite designs [@Box1951-ji] via functions `bbd()` and `ccd()`, respectively, where the minimum required input is the number of factors. Another form of factorial design is the *saturated designs* where higher-order interaction effects of treatment factors are typically confounded with the main effects. Plankett-Burnman designs [@Plackett1946-ly] are a type of saturated design that can be generated by the function `pb()` in the `r CRANpkg("FrF2")` package, where the user provides the number of experimental runs and the number of treatment factors.

@@

The `r CRANpkg("agricolae")` package [@agricolae] is the prime example of constructing designs based on a set of so-called "recipe functions", where each function corresponds to a single class of experimental design. For example, `design.crd()`, `design.rcbd()` and `design.split()` construct completely randomized, randomized complete block and split-plot designs, respectively. Users typically supply treatment labels (or the number of treatments in the case of `design.split()`) and the number of replications as arguments for these functions. The output is a list with one element corresponding to a `data.frame` that contains the design in a table such that the row corresponds to the experimental run and the columns correspond to the experimental variables (we refer to this format simply as "table format" henceforth).

The use of recipe functions is not limited to classical experimental designs; the `r CRANpkg("AlgDesign")` [@AlgDesign] package offers three primary functions for generating optimal designs: `optBlock()`, `optFederov()` and `optMonteCarlo()`. In general, these functions require data and formulas in terms of the supplied data variables with the choice of the criterion (e.g. the D-criterion), with the output as a list with one element corresponding to the design in a table format. The difference between these functions lies in the underlying search strategy for optimal designs, and the name of the function is a surrogate for the search algorithm.

Computer experiments, which generally involve space-filling designs, are implemented in packages such as `r CRANpkg("lhs")` [@lhs] and `r CRANpkg("DiceDesign")` [@DiceDesign]. For the `r CRANpkg("lhs")` package, functions such as `randomLHS()`, `optimumLHS()`, and `maximinLHS()`, require users to specify the sample size ($n$) and the number of variables ($p$), then it generates a Latin hypercube sample [@McKay1979-aw], based on different optimization schemes (in this case, random, S optimal and maxmin criteria, respectively; see package documentation for more details). Similarly for `r CRANpkg("DiceDesign")`, there is a comprehensive list of space-filling designs such as `dmaxDesign()`, `lhsDesign()`, and `wspDesign()` with input of $n$ and $p$ as before (among additional parameters for some) that implement algorithms that either maximize the entropy [@Shewry1987-jn], produce a random Latin hypercube sample, or use the WSP algorithm [@Santiago2012-ui], respectively. These designs output an $n \times p$ matrix with values between 0 and 1. Again, these functions employ a recipe style, in which each function has a name that corresponds to a certain search strategy to generate the experimental design.

@@

Some functions in the DoE packages require the input of existing experimental designs to produce new designs. For example, the `r CRANpkg("DoE.base")` package contains some experimental functions, `cross.design()` and `param.design()`, to combine designs with the former taking a Cartesian product of the input designs while the latter uses a Taguchi style [@Taguchi1986-ut] to aggregate the designs with the inner and outer arrays. The `r CRANpkg("lhs")` package contains a function that `augmentLHS()` to add additional samples to the existing Latin hypercube sample.

Another class of augmenting design is *sequential design* (also called *adaptive sampling*), which is best represented by `r CRANpkg("tgp")` [@tgp]. This requires prior information that is used to inform the next experimental design using `tgp.design()` and `dopt.gp()`. The user is required to supply candidate samples to subsample from and a model or a prior experimental design. Follow-up experiments, which can also be classified as sequential designs, is implemented by `r CRANpkg("BsMD")` [@BsMD] using a model-discriminant approach using `MD()`.

@@

Through the exploratory data analysis of the three data sources (package download logs, package metadata and CRAN task views) outlined in `r secref("data", "Data")`, we observed in `r secref("eda", "Exploratory data analysis")` that the the total download of DoE packages is concentrated only on a handful of R-packages although these represent a diverse set in comparison with other CRAN task views. Furthermore, the data suggest that experimental design is the least collaborative field.

@@

The interface design (discussed in `r secref("design", "Interface design")`) reveals that the most widely used DoE packages generally have functions that 1) focus on certain aspects of experimental design (e.g. factorial structure or augmenting design), 2) are a recipe format (i.e. the name of the function is a surrogate to a single class of the design or optimal search algorithm) and 3) context is often a second thought -- many inputs a single integer corresponding to the number of factors, levels, or experimental runs (or sample size). The function will often assign pseudo-factor names or there is an optional argument to input a character vector that corresponds to the factor names. These interface designs require users to have processed the experiment in statistical terms (often stripping the experimental context away) and simultaneously, users must choose the generation mechanism by selecting an appropriate function. Arguably, the current dominant interface designs are not aligned with the way practitioners cognitively design their experiments. Often, the critical part of designing an experiment is to understand the experimental structure, and the experimental context can govern or guide the choice of algorithm to allocate treatments. In addition, the nature of recipe functions can obscure the understanding and relation of designs (e.g., how do you go from an unstructured factorial design to split plot design?). Each new method for generating an experimental design appears to correspond to a completely new function, and intermediate results are often not easily accessible. These factors may contribute to why developers often work in silos in the field of experimental design.

A consistent cognitive interface design that leverages existing developments and can be easily extended to new methods will conceivably be of great help to practitioners. Some efforts to this end are seen in `r CRANpkg("DoE.wrapper")` [@DoE.wrapper], which contains wrapper recipe functions for other DoE packages such as `r CRANpkg("lhs")`, `r CRANpkg("AlgDesign")` and `r CRANpkg("FrF2")` (see Figure `r ref("fig:plot-doe-network")`), and is also the subject of the developmental package `r CRANpkg("edibble")` [@edibble]. Undoubtedly, no single developer or package can cater to all experimental designs; therefore, any unifying interface should consider how other developers can contribute or add their methods. Future research could benefit from further exploratory data analysis, expanding the study to beyond R-packages, and discussing other aspects of interface designs.

@@

This paper uses `r CRANpkg("targets")` [@targets] and `r CRANpkg("renv")` [@renv] for reproducibility, `r CRANpkg("knitr")` [@knitr] and `r CRANpkg("rmarkdown")` [@rmarkdown] for creating reproducible documents, `r CRANpkg("ggplot2")` [@ggplot2], `r CRANpkg("ggraph")` [@ggraph], `r CRANpkg("ggwordcloud")` [@ggwordcloud] and `colorspace` [@colorspace] for visualization, `r CRANpkg("kableExtra")` [@kableExtra] for customizing the table in the Supplementary Material and `r CRANpkg("tidyverse")` [@Wickham2019-mj], `r CRANpkg("tidytext")` [@tidytext], `r CRANpkg("pluralize")` [@pluralize], and `r CRANpkg("ineq")` [@ineq] for data processing and manipulation, and `r CRANpkg("cranlogs")` [@cranlogs] and `r CRANpkg("ctv")` [@ctv] for extracting data. All codes used to reproduce this paper are found in `r url("https://github.com/emitanaka/paper-DoE-review")`.

### 2022-12-14 paper/arxiv/arxiv.Rmd
reference  setups for specific outlets to accommodate differences in non-numbered/numbered sections and output types (PDF/HTML)

---
authors:
  - name: Emi Tanaka
    thanks: Corresponding author
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Dewi Amaliah
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
header-includes: |
  \usepackage{colortbl}
  \usepackage{bera}
  \usepackage{xcolor}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \newcommand{\ctv}[1]{\href{http://CRAN.R-project.org/view=#1}{\emph{#1}}}
bibliography: paper.bib
biblio-style: unsrt
#preamble: >
output: 
  rticles::arxiv_article:
    keep_tex: true
---

@@

```{r child=paper}
```

>>>>

---
authors:
  - name: Emi Tanaka
    thanks: Corresponding author
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Dewi Amaliah
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
header-includes: |
  \usepackage{colortbl}
  \usepackage{bera}
  \usepackage{xcolor}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \newcommand{\ctv}[1]{\href{http://CRAN.R-project.org/view=#1}{\emph{#1}}}
  \newcommand{\CRANpkg}[1]{\href{https://cran.r-project.org/web/packages/#1/index.html}{\texttt{#1}}}
bibliography: paper.bib
biblio-style: unsrt
#preamble: >
output: 
  rticles::arxiv_article:
    keep_tex: true
---

@@

```{r, echo= FALSE, results="asis"}
params <- list()
params$ref_setup_fn <- "paper/arxiv/ref-setup.R"
cat(knitr::knit_child(paper, envir = environment(), quiet = TRUE))
```
